---
title: "Situational Pitching to Contact Analysis"
author: "Judah Axelrod"
---

Setup
-Import libraries and multispread.R function; 
-Connect to Statcast MySQL database with pitch-level data from 2008-2019
-Import Fangraphs data on qualified pitcher ERA, FIP, and xFIP from 2015-2019
(Setup Code partially hidden)

```{r echo=FALSE,message = FALSE,warning=FALSE}
library(tidyverse)
library(RMySQL)
library(ggpmisc)
library(gridExtra)

```

1. Query database to obtain raw data (all pitches thrown from 2015-2019)
2. Calculate K% and BB% for each pitcher and year, separately with runners on and bases empty
3. Calculate K_diff and BB_diff (difference in K% or BB% for runners on vs. bases empty)
4. For each pitcher and year, obtain K_diff and B_diff for both the current and previous year

```{r message=FALSE,warning=FALSE}
k_bb_annual_diffs <- function(sc_year){
    query <- paste("
    select distinct a.game_year, a.game_pk, a.pitcher, a.batter, a.at_bat_number, 
      a.on_1b, a.on_2b, a.on_3b, a.outs_when_up, a.description, a.events,
      a.bb_type, b.name_lahman
    from sc_all as a left join playerids as b
    on a.pitcher = b.mlbcode
    where game_year =",sc_year,"and events != 'null';")
    pitchers <- dbGetQuery(db, query)
    


    c0_pitchers <- pitchers %>%
      mutate(runners_on = case_when(!is.na(on_1b) | !is.na(on_2b) | !is.na(on_3b) ~ 1,
                                    TRUE ~ 0),
             scoring_position = case_when(!is.na(on_2b) | !is.na(on_3b) ~ 1,
                                          TRUE ~ 0))
    
    strikeouts_walks <- c0_pitchers %>%
      filter(events %in% c("strikeout","strikeout_double_play","walk")) %>%
      mutate(outcome = case_when(events == "walk" ~ "Walks",
                                 TRUE ~ "Strikeouts")) %>%
      count(game_year, name_lahman, outcome, runners_on)
    
    
    strikeout_walk_pct <- c0_pitchers %>%
      count(name_lahman, runners_on) %>%
      rename(total_events = n) %>%
      group_by(name_lahman) %>%
      mutate(TBF = sum(total_events)) %>%
      filter(TBF >= 100 & !is.na(name_lahman) & name_lahman != "NA NA") %>%
      inner_join(strikeouts_walks, by = c("name_lahman","runners_on")) %>%
      spread(key = "outcome", value = "n") %>%
      multi_spread(key = "runners_on", value = c("total_events","Strikeouts","Walks")) %>%
      mutate(k_pct_empty = `0_Strikeouts` / `0_total_events`,
             k_pct_on = `1_Strikeouts` / `1_total_events`,
             bb_pct_empty = `0_Walks` / `0_total_events`,
             bb_pct_on = `1_Walks` / `1_total_events`,
             k_diff = k_pct_on - k_pct_empty,
             bb_diff = bb_pct_on - bb_pct_empty,
             TBF = `0_total_events`+`1_total_events`) %>%
      arrange(desc(k_diff))
    
    strikeout_walk_pct %>%
      select(game_year, name_lahman, TBF, starts_with("k"), starts_with("bb"))
    
    }


k_bb_diffs15 <- k_bb_annual_diffs(2015)
k_bb_diffs16 <- k_bb_annual_diffs(2016)
k_bb_diffs17 <- k_bb_annual_diffs(2017)
k_bb_diffs18 <- k_bb_annual_diffs(2018)
k_bb_diffs19 <- k_bb_annual_diffs(2019)
                         
k_bb_diffs_5yr <- rbind(k_bb_diffs15,k_bb_diffs16,k_bb_diffs17,k_bb_diffs18,k_bb_diffs19)


k_bb_diffs_lag <- k_bb_diffs_5yr %>%
  arrange(name_lahman, game_year) %>%
  mutate(k_diff = 100*k_diff, bb_diff = 100*bb_diff,
        lag_k_diff = case_when(name_lahman == lag(name_lahman) ~ lag(k_diff)),
         lag_bb_diff = case_when(name_lahman == lag(name_lahman) ~ lag(bb_diff))) %>%
  filter(!is.na(lag_bb_diff) & !is.na(lag_k_diff))

```

Plot Year 1 K_diff and BB_diff against Year 2 K_diff and BB_diff

```{r message=FALSE,warning=FALSE}
p1 = ggplot(k_bb_diffs_lag, aes(x = lag_k_diff, y = k_diff)) + geom_point(size=0.75,color='darkred') + geom_smooth(method='lm', formula= y~x, color='black') +
  labs(x=expression('Year 1 K%'[diff]), y=expression('Year 2 K%'[diff])) + theme_bw()

p2 = ggplot(k_bb_diffs_lag, aes(x = lag_bb_diff, y = bb_diff)) + geom_point(size=0.7,color='dodgerblue') + geom_smooth(method='lm', formula= y~x, color='black') +
  labs(x=expression('Year 1 BB%'[diff]), y=expression('Year 2 BB%'[diff])) + theme_bw()

grid.arrange(p1,p2,ncol=2)

```               


Look for all pitchers who had K_diff or BB_diff > 2.5% to examine further, and obtain their ERA, FIP, and xFIP for 2015-2019

```{r message=FALSE,warning=FALSE}

k_bb_diffs_agg <- k_bb_diffs_5yr %>%
  group_by(name_lahman) %>%
  summarize(N=n(), avg_k_diff = 100*mean(k_diff,na.rm=TRUE), 
            avg_bb_diff = 100*mean(bb_diff,na.rm=TRUE)) %>%
  filter(N>2 & (avg_k_diff > 2.5 | avg_bb_diff > 2.5)) %>%
  arrange(desc(avg_k_diff))
  
c0_era_fip <- 
  inner_join(k_bb_diffs_5yr,era_fip,by=c('name_lahman'='Name','game_year'='Season'))

c0_era_fip_lim <- 
  inner_join(k_bb_diffs_5yr,era_fip,by=c('name_lahman'='Name','game_year'='Season')) %>%
  inner_join(k_bb_diffs_agg, by=c('name_lahman'))
  
era_fip_lag <- function(data){
  data %>%
    arrange(name_lahman, game_year) %>%
    mutate(lag_ERA = case_when(name_lahman == lag(name_lahman) ~ lag(ERA)),
           lag_FIP = case_when(name_lahman == lag(name_lahman) ~ lag(FIP)),
           lag_xFIP = case_when(name_lahman == lag(name_lahman) ~ lag(xFIP))) %>%
    filter(!is.na(lag_ERA) & !is.na(lag_FIP)) %>%
    select(name_lahman, game_year, ERA, FIP, xFIP, lag_ERA, lag_FIP, lag_xFIP)
} 
c0_era_fip_lag <- era_fip_lag(c0_era_fip)
c0_era_fip_lag_lim <- era_fip_lag(c0_era_fip_lim)

```

Plot Year 1 ERA, FIP, and xFIP against Year 2 for both All Qualified Pitchers and the subset found above

```{r message=FALSE,warning=FALSE}
p3 = ggplot(c0_era_fip_lag) + geom_point(aes(x=lag_ERA, y=ERA),color='darkred',size=0.5) + 
  geom_smooth(aes(x=lag_ERA,y=ERA,color='ERA'),method='lm',se=F) +
  geom_point(aes(x=lag_FIP, y=ERA),color='navyblue',size=0.5) +
  geom_smooth(aes(x=lag_FIP, y=ERA,color='FIP'),method='lm',se=F) +
  geom_point(aes(x=lag_xFIP, y=ERA),color='darkgrey',size=0.5,alpha=0.5) +
  geom_smooth(aes(x=lag_xFIP, y=ERA,color='xFIP'),method='lm',se=F) + theme_bw() +
  scale_color_manual(name='Legend',values=c('darkred','navyblue','darkgrey')) +
  guides(color=F) +
  labs(x='Year 1 ERA, FIP, xFIP',y='Year 2 ERA') +
  ggtitle('All Qualified Pitchers') + theme(plot.title = element_text(face='bold',hjust = 0.5))
p4 = ggplot(c0_era_fip_lag_lim) + geom_point(aes(x=lag_ERA, y=ERA),color='darkred',size=0.5) + 
  geom_smooth(aes(x=lag_ERA,y=ERA,color='ERA'),method='lm',se=F) +
  geom_point(aes(x=lag_FIP, y=ERA),color='navyblue',size=0.5) +
  geom_smooth(aes(x=lag_FIP, y=ERA,color='FIP'),method='lm',se=F) + 
  geom_point(aes(x=lag_xFIP, y=ERA),color='darkgrey',size=0.5,alpha=0.7) +
  geom_smooth(aes(x=lag_xFIP, y=ERA,color='xFIP'),method='lm',se=F) + theme_bw() +
  scale_color_manual(name='Legend',values=c('darkred','navyblue','darkgrey')) +
  labs(x='Year 1 ERA, FIP, xFIP',y='Year 2 ERA') +
  ggtitle('Situational Pitchers') + theme(legend.position = c(-0.05,0.8),legend.background=element_rect(linetype='solid',color='black'),
                                            legend.title = element_text(face='bold'),plot.title = element_text(face='bold',hjust = 0.5))
grid.arrange(p3,p4,ncol=2)

```
